---
title: "Exam 1 - Rosairi Estevez"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
  word_document:
    toc: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

### Rosairi Estevez

### PSY V0500 Statistical Methods in Psychology

### March 14, 2024

### Exam 1

```{r}
library(ggplot2)
```

```{r}
library(tidyverse)
```

```{r}
attach(brfss22)
```

*The questions are worth 120 points. You have 120 minutes to do the
exam, one point per minute.* *All answers should submitted
electronically. Please submit all relevant computer files as a Slack
message to me (perhaps just a message with GitHub link). I prefer .Rmd
files **along with knit output,** md or html is fine. Graphs should be
'export'. Or you can save output and code as pdf or rtf if you must.*
*You may refer to your books, notes, calculator, computer, etc. The exam
is "open book."* *However, you must not refer to anyone else, either in
person or electronically, during the exam time. For instance, since
these exam questions are newly created, posting questions or copying
answers on Discord or WhatsApp groups or online homework helping sites
or forums (such as Chegg, Yahoo answers or others) is a violation. Don't
upload to public GitHub site until end of exam.* *You must do all work
on your own. Cheating is harshly penalized.* *Good luck. Stay cool.*

The first questions do not require any work in R (although you might
find it convenient, I'm not stopping you) -- I will provide some summary
data and you can construct hypothesis tests on your own. They're meant
to test your understanding of the stats, apart from coding. Subsequent
questions require R coding.

We will look at marijuana usage since this is an important research
topic of many of the faculty here. While BRFSS asks how many days (of
last 30) a person used that, I recode into just zero (for zero days) and
one (for one or more days), `use_marijan`.

1.  (25 points) I've run crosstabs (just use these data as provided) of
    marijuana usage with Age (as in lab, I give the midpoint of the age
    range). It's interesting to explore usage rates by age to see how
    that varies.

Combine some or all of these to form one or more hypothesis tests about
differences. Explain carefully the question(s) and answer(s) of the
hypothesis test(s). I expect that different people may choose different
levels of significance. Please provide estimate, standard error, t-stat
and a p-value for each hypothesis test and a confidence interval.
**Write a paragraph** explaining each test (carefully noting what is the
null hypothesis) and explaining the results of that test along with what
assumptions are needed for the test to be valid.

```{r eval=FALSE}
xtabs(~ use_marijan + Age_midpt)
                         Age_midpt
use_marijan   21   27   32   37   42   47   52   57   62   67   72   77   82
          0 3737 3117 3675 4530 5069 4949 6273 6960 8520 9699 9275 7154 7963
          1 1311  990 1098 1050 1073  828  820  936 1111 1066  758  311  105
```

`Ho (Null Hypothesis): Group A = Group B` There will be no differences
between proportions of 21 - 27 year olds and 32 - 37 year olds in terms
of marijuana use, they will be the same.

`Ha (Alt Hypothesis): Group A  ≠ Group B` There will be differences
between proportions of 21 - 27 years olds and 32-37 year olds in terms
of marijuana use, they will not be the same.

sig. level = 0.05 95%

Group A = Age 21 - 27 of marijuana users n = 9,155

marijuana users = 2,301

p1 = 0.251

Group B = Age 32 - 37 of marijuana users n = 10,353

marijuana users = 2,148

p2 = 0.207

`point estimate` = 0.251 - 0.207 = 0.044

`standard error` = 0.006

`confidence interval` = (0.032, 0.055)

`test statistic` = 7.287

`p val` = 3.178e-13

My hypothesis test regarded ages 21 - 27 (Group A) versus ages 32 - 37
(Group B) and their marijuana uses. Testing whether the proportions of
each group would either be the same (Ho: Group A = Group B) or would
differ (Ha: Group A ≠ Group B). We *reject null hypothesis* because the
t-stat which is 7.287 does not fall within the 95% confidence intervals
proportion (0.032, 0.055) nor in the 95% region of acceptance (-1.96,
1.96). Meaning there is a significant difference in the proportions
between the age groups and marijuana use.

```{r}
rm(list = ls())
prop.test(x=c(2301, 2148), n=c(9155, 10353), alternative="two.sided", conf.level =0.95, correct = FALSE)
```

2.  (25 points) Continue to explore marijuana usage, now looking at
    educational levels. Form a hypothesis test about differences in
    rates (*not* levels). Explain carefully the question and answer of
    the hypothesis test. I expect that different people may choose
    different levels of significance. Please provide estimate, standard
    error, t-stat and a p-value for the hypothesis test and a confidence
    interval. **Write a short paragraph** explaining the test (carefully
    noting what is the null hypothesis) and explaining the results of
    that test along with what assumptions are needed for the test to be
    valid.

```{r eval=FALSE}
educ_shortlabel <- fct_recode(EDUCA, "never" = "Never attended school or only kindergarten",
                    "elementary" = "Grades 1 through 8 (Elementary)", 
                    "some hs" = "Grades 9 through 11 (Some high school)",
                    "hs" = "Grade 12 or GED (High school graduate)",
                    "some coll" = "College 1 year to 3 years (Some college or technical school)",
                    "degree" = "College 4 years or more (College graduate)",
                    "Refused" = "Refused")

xtabs(~ use_marijan + educ_shortlabel)
                         educ_shortlabel
use_marijan never elementary some hs    hs some coll  degree  Refused
          0    95       1472    2642 20004     22747   34993     257
          1     8        104     587  3297      3555    3958      16

```

`Ho (Null): Group 1 = Group 2` The rate of marijuana use between
individuals who have college degrees versus individuals who only went to
high school will be the same/no difference.

`Ha (Alt): Group 1 ≠ Group 2` The rate of marijuana use between
individuals who have college degrees or have only gone to high school
will NOT be the same.

sig. level = 0.05 95%

Group 1 (Degree - marijuana) n = 38951 p1 = 3958/38951 = 0.102

Group 2 (HS - marijuana) n = 20004 p2 = 3297/20004 = 0.165

`point estimate` = 0.063

`standard error` = 0.003

`confidence interval` = (-0.069, -0.057)

`test statistic` = -22.118

`p val` = 0

My hypothesis test compared marijuana use rates between individuals with
degrees versus individuals who only completed HS. The null hypothesis is
that there will be no difference in use rates between the two groups
(Ho: Group 1 = Group 2), and my alternative hypothesis is that there
will be a difference (Ha: Group 1 ≠ Group 2) We *reject the null
hypothesis* due to the t-stat = -22.118 not being within the 95% region
of acceptance (-1.96, 1.96) and falling outside the critical values.

```{r}
rm(list = ls())
prop.test(x=c(3958, 3297), n=c(38951, 20004), alternative="two.sided", conf.level =0.95, correct = FALSE)
```

3.  (30 points) Now create some interesting graphs with ggplot on either
    of the previous two topics: how marijuana use correlates with age or
    education. Explain what conclusion you would draw from the graphs.
    In what ways do the graphs illustrate or complicate the statistical
    tests?

*Conclusions I can draw from the graphs are that marijuana use was
definetly higher in individuals who got degrees, and then close runner
ups being individuals who had some college and high school.*

The graphs illustrate the statistical test in some way where it clearly
shows that there are significant differences in rates of marijuana use
between the different educational levels. It seems that after hs is when
the numbers drastically go up, so it can be a link with academic stress
and so on.

```{r}
ggplot(brfss22, aes(x = educ_shortlabel, fill = MARIJAN1)) + geom_bar()
```

```{r}
p <- ggplot(data = brfss22,
            mapping = aes(x = MARIJAN1,
                          y = educ_shortlabel))
```

```{r}
p2 <- ggplot(data = brfss22,
            mapping = aes(x = educ_shortlabel,
                          y = MARIJAN1))
```

```{r}
p + geom_point()
```

```{r}
p2 + geom_point()
```

```{r}
p + geom_point(color = "purple") + 
  geom_smooth(method = "loess") +
  scale_x_log10()
```

```{r}
ggplot(brfss22, aes(x = MARIJAN1, fill = educ_shortlabel)) + geom_bar(position = "fill") + scale_fill_viridis_d(option="rocket")
```

```{r}
p + geom_line()
```

4.  (40 points) Now expand the analysis to look at other correlates with
    marijuana use.

-   Choose a subgroup of the sample to consider and provide summary
    statistics of that subgroup. Explain why this subgroup is
    interesting.
-   Form a hypothesis test about an interesting variable, explore
    whether your chosen subgroup differs from the rest of sample. Please
    provide both a p-value for the hypothesis test and a confidence
    interval. Write a short paragraph explaining the test (carefully
    noting what is the null hypothesis) and explaining the results of
    that test.
-   Using a k-nn classifier, can you find relevant information to
    predict marijuana use? How good is the classifier? Discuss.
-   Can you explain some other interesting information about this data?
    Some interesting crosstabs? Some good hypothesis tests? Impress me.

subgroup = adverse childhood experiences drugs **This subgroup is
interesting since often experiences of physical abuse have been
correlated towards behaviors of substance abuse and so on. It'd be
interesting if people with adverse childhood experiences of punching
either once or more than once have differences in marijuana uses.**

`Ho = Individuals with and without ACEdrugs experiences will no difference in proportions of marijuana use`
`Ha = Individuals with and without ACEdrugs experiences will have difference in proportions of marijuana use`

group 1 - acepunch once + marijuana use

n =

group 2 - no acepunch more than once + marijuana use

n =

```{r}
xtabs(~ brfss22$MARIJAN1 + brfss22$ACEPUNCH)
```

```{r}
prop.table(table(brfss22$MARIJAN1, brfss22$ACEPUNCH))
```

```{r}
brfss_ACE$ACEPUNCH_recode <- fct_recode(brfss_ACE$ACEPUNCH, 
                                        "0" = "Adverse Childhood Exper, never: How often did your parents or adults in your home ever slap, hit, kick, punch or beat each other up",
                                "0.5" = "once", 
                                "1" = "more than once",
                                NULL = "dont know not sure",
                                NULL = "refused"
)

brfss_ACE$ACEPUNCH_recode <- as.numeric(levels(brfss_ACE$ACEPUNCH_recode))[brfss_ACE$ACEPUNCH_recode]

```

```{r}
summary(MARIJAN1)
```

```{r}
MARIJAN1_recode <- fct_recode(brfss22$MARIJAN1, 
                                        "1" = "Yes, Adverse Childhood Exper, lived with someone who used illegal street drugs or who abused prescription medications",
                                "0" = "No", 
                                NULL = "dont know not sure",
                                NULL = "refused"
                            
)

summary(brfss_ACE$ACEDRUGS_recode)
```

```{r}
brfss_ACE <- subset(brfss22, select_ACE)
```

```{r}
select_ACE <- (ACEdidntask == 0)
```

```{r}
ACEdidntask <- (as.numeric(is.na(brfss22$ACEDEPRS)) + 
                        as.numeric(is.na(brfss22$ACEDRINK)) +
                        as.numeric(is.na(brfss22$ACEDRUGS)) +
                        as.numeric(is.na(brfss22$ACEPRISN)) +
                        as.numeric(is.na(brfss22$ACEDIVRC)) +
                        as.numeric(is.na(brfss22$ACEPUNCH)) +
                        as.numeric(is.na(brfss22$ACEHURT1)) +
                        as.numeric(is.na(brfss22$ACESWEAR)) +
                        as.numeric(is.na(brfss22$ACETOUCH)) )
```

```{r}
standardize_varb_to01 <- function(X_in) {
  (X_in - min(X_in, na.rm = TRUE))/( max(X_in, na.rm = TRUE) - min(X_in, na.rm = TRUE)  )
```

```{r}
X1 <- standardize_varb_to01(brfss_ACE$MARIJAN1)
X2 <- brfss_ACE$ACEPUNCH_recode

```
